---
title: "Counting at the Microscope: The count_app() Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Counting at the Microscope: The count_app() Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

`count_app()` is the primary interface in pcountr. It launches a keystroke-driven
Shiny application designed for use at the microscope: type an ID and a preservation code, press
Enter or Space, and the count is recorded. Live running totals, continuous autosave, and
a resume-from-YAML path make it a practical replacement for the PCount DOS
counting loop.

pcountr is not limited to pollen. The dictionary format, input tokens, and YAML
output are proxy-agnostic — diatom, charcoal morphotype, phytolith, or any other
assemblage counted by traversing a slide follows the same workflow unchanged.

---

## What you need before you start

Three things:

1. **A dictionary** — either a PCount `.DIC` file or a CSV in the pcountr
   template format. The dictionary maps taxon codes to names and sum groups.
   See `inst/extdata/dictionary_template.csv` for the column layout.

2. **Spike parameters** — the number of spike tablets (or volume/mass) added
   per sample, and the concentration of your spike standard.

3. **A save folder** — a directory where YAML count files will be written. One
   file is created per sample.

If you are transitioning from PCount, your existing `.DIC` file works directly —
load it with `read_dic()` or point the app setup screen at it.

---

## Launching the app

```{r launch, eval = FALSE}
library(pcountr)
count_app()
```

The app opens in your default browser. It requires the `shiny`, `DT`, and
`shinyFiles` packages; install them if prompted.

---

## Setup screen

The first screen collects instrument and sample metadata before counting begins.

| Field | Notes |
|---|---|
| **Dictionary** | Browse to a `.DIC` or `.csv` dictionary file |
| **Sample name** | Free text, e.g. `KF24sh#001`; accepts any string |
| **Depth top / bottom** | Interval in cm; leave blank if unknown |
| **Age top / bottom** | Calibrated years BP; leave blank if unknown |
| **Calculate concentration?** | `Yes, using spikes` (Stockmarr equation), `Yes, volumetrically` (ΣP / sample quantity), or `No` |
| **Spike tablets** | Number of tablets added — shown only when method is `spike` |
| **Spike density** | Microspheres per unit of spike — shown only when method is `spike` |
| **Spike units** | `tablets`, `ml`, or `g` — shown only when method is `spike` |
| **Use preservation codes?** | `Yes` (grain tokens require a preservation digit, e.g. `B1`) or `No` (code only, e.g. `B`) |
| **Sample quantity** | Volume or mass of sediment processed |
| **Sample units** | `ml` or `g` |
| **Save path** | Full path for this sample's YAML file |

Click **Start Counting** when the fields are complete. The save path is locked
once counting begins; all other metadata fields remain editable on the
*Sample Info* tab during the count.

---

## Input syntax

Type any of the following in the input field and press **Enter** or **Space** to submit:

| Token | Meaning |
|---|---|
| `B1`, `I80`, `A1` | **With preservation codes (default):** taxon code + base preservation digit + optional modifiers |
| `B10` | Half-grain modifier: code `B`, preservation `1`, modifier `0` (weight 0.5) |
| `B19` | Hidden grain: code `B`, preservation `1`, modifier `9` |
| `B9`, `B0` | Modifier alone — a grain recorded as hidden, or as a half-grain, with no base preservation state. The base digit is optional |
| `B`, `I`, `A` | **Without preservation codes:** code only, when "Use preservation codes?" is set to No |
| `.` | One tracer microsphere (spike) — only when concentration method is `spike`. Pressing `.` on an empty field submits immediately (no Enter needed) |
| `/label/` | Traverse marker — stored verbatim in the grain sequence |
| `[text]` | Inline remark — stored verbatim |

Modifiers: `9` = hidden, `0` = half-grain (weight 0.5).

Digits `1`–`8` are preservation states, and **what they mean is up to you.** The
built-in defaults are `1` well-preserved, `2` corroded, `6` crumpled, `8` broken
(after Cushing, 1967); `3`, `4`, `5`, `7` carry no label. To use your own
damage-state taxonomy, pass a named vector to `pollen_site(preservation = )`;
every summary and label follows from it.

Unknown or malformed tokens trigger a prominent red alert and an audio beep.
The grain is **not recorded** — correct and re-enter.

Use the **New Slide** button when moving to a new slide; you will be prompted
for a slide ID, which is stored as a traverse marker.

---

## Live metrics

The header panel updates after every entry:

- **Σ** — total grains counted (all groups)
- **ΣP** — group-filtered sum (e.g., terrestrial pollen types only)
- **Traverses** — running traverse count
- **Spike** — total spike grains recorded
- **Concentration** — grains/g or grains/cm³; method depends on the setup choice
  (spike: Stockmarr equation; volumetric: ΣP / sample quantity; none: always `NA`).
  Shows `NA` until all required inputs are present
- **PAR** — pollen accumulation rate; shows `NA` until depth and age intervals
  are both present

---

## Autosave and resume

The YAML file is written continuously — grain entries save within ~300 ms, and
all other actions (traverses, remarks, undo, metadata changes) save immediately.
Closing the browser or losing power risks losing at most a handful of the most
recent rapid keystrokes. The *Done / Save* button is a confirmation step only.

To resume a previous count, click **Open Existing Count** on the setup screen
and navigate to the YAML file. Grain order, traverses, remarks, spike total, and
all metadata are restored exactly as they were.

After finishing a sample, **New Sample** carries over instrument metadata
(dictionary, spike parameters, sample units) and suggests an incremented save
path, ready for the next slide.

---

## After counting: loading YAML files

Once your count session is complete, load the saved YAML files with `read_site()`.
Depths, ages, and spike parameters embedded during counting are read automatically
— no separate metadata sheet is needed. You do need to supply the dictionary:
`read_site()` requires it to assign taxa to sum groups (A/B/F/…) for all
downstream calculations. Pass the same CSV dictionary you used in the app.

```{r load-yaml, eval = FALSE}
site <- read_site("path/to/your/yaml/folder",
                  dic = "path/to/your/dictionary.csv")
print(site)
```

If you have a mix of new YAML counts and legacy `.CNT` files in the same folder,
`read_site()` loads both transparently.

The examples below use the **Fake Lake** dataset bundled with pcountr — 20
simulated samples loaded from `.CNT` files — to demonstrate the post-counting
workflow with runnable code. The output is identical to what you would get from
a folder of YAML files produced by `count_app()`.

```{r fake-lake}
library(pcountr)

cnt_dir   <- system.file("extdata", "fake_lake", package = "pcountr")
meta_path <- system.file("extdata", "fake_lake", "metadata_FL.csv",
                         package = "pcountr")
dic_path  <- system.file("extdata", "fake_lake", "ECG.csv",
                         package = "pcountr")

site <- read_site(cnt_dir, metadata = meta_path, dic = dic_path)
print(site)
```

---

## Correcting metadata after counting

If you discover a depth error, a wrong age, or a sample title that needs
updating, you can correct it directly in a CSV without reopening the app or
editing the YAML by hand.

`extract_metadata()` exports all sample metadata to a data frame or CSV file.
Edit the CSV in any spreadsheet editor — only the cells you change matter — and
pass it back to `apply_metadata()`:

```{r metadata-roundtrip, eval = FALSE}
site <- read_site("path/to/your/yaml/folder",
                  dic = "path/to/your/dictionary.csv")

# Export metadata for review
extract_metadata(site, file = "metadata_check.csv")

# … open metadata_check.csv, correct any errors, save …

# Apply edits and write updated YAMLs
site <- apply_metadata(site, "metadata_check.csv")
```

`apply_metadata()` matches each CSV row to a sample by `source_file` (full path
stored in the YAML — preferred) or by `sample_name` as a fallback. Every
non-NA cell is applied; columns you leave blank or NA are not overwritten. Each
modified sample is written back to its source YAML file. To preview changes in
memory without writing to disk, pass `write = FALSE`:

```{r metadata-preview, eval = FALSE}
site_preview <- apply_metadata(site, "metadata_check.csv", write = FALSE)
```

All fields exported by `extract_metadata()` are editable this way: `depth_top`,
`depth_bottom`, `age_top`, `age_bottom`, `sample_name`, `sample_quantity`,
`units`, `spike_tablets`, `spike_density`, `spike_units`, `conc_method`, and
`title`. The `source_file` column identifies each YAML file and is never
overwritten.

### Backfilling `spike_units` in files saved before v0.5.4

YAML files created by `count_app()` before v0.5.4 do not contain a `spike_units`
field due to a bug in the autosave path. To backfill the field across an existing
folder of files:

```{r backfill-spike-units, eval = FALSE}
site <- read_site("path/to/yaml/folder",
                  dic = "path/to/your/dictionary.csv")
extract_metadata(site, file = "meta_fix.csv")
# Open meta_fix.csv, fill the spike_units column for each sample, then:
site <- apply_metadata(site, "meta_fix.csv")
```

---

## How many grains should I count?

`rarefaction()` estimates this from counts you have already made. For each
sample it builds the rarefaction curve, extrapolates the richness asymptote
(`Smax`) with a Michaelis–Menten model, and reports the grain counts needed to
recover 70%, 80%, and 90% of it.

```r
r <- rarefaction(site)
r
```

The `%Smax` column is what your count actually recovered; the three tier columns
are what further effort would buy. The final row gives a site-level target — the
90th percentile across samples — so one number covers most of the core.

**Why not simply count until the curve flattens?** Because a curve built from
one sample flattens at that sample's own richness by construction: drawing all
your grains always recovers every taxon you found. Judging a count against its
own richness always concludes that it was sufficient. The asymptote has to be
extrapolated *past* what you counted, which is what the model does — see
`DESIGN.md` §12.

**No verdict is reported, deliberately.** Adequacy depends on what you are doing
with the data. Lesven et al. (2026) found ~250 grains sufficient to reconstruct
dominant vegetation assemblages, recovering ~70% of richness, but recommend
~1000 grains for biodiversity assessment or detecting rare taxa (85–95% of
richness). Iglesias et al. (2017) likewise report their optima as consistent with
the established convention of 300 grains for standard studies and 500–1000 where
rare pollen types are of interest. Pick the tier that fits your question.

Two things worth knowing when you read the output. `Smax` is a model
extrapolation, not an observation — Lesven et al. found no curve reaching a true
asymptote even at 1000 grains, and fits from ~300-grain counts are still on the
steep part of the curve, where the asymptote is poorly constrained. Treat the
targets as provisional and read `Smax` and `k` alongside them. And evenness
matters alongside richness: Iglesias et al. (2017) note that the count needed
depends on both, since a sample dominated by one taxon spends much of its count
on that taxon.

Samples too small to extrapolate (fewer than 30 grains or 4 taxa), or whose fit
fails a sanity check, print `--` rather than a fabricated number, and are left
out of the site target.

---

## Finding your remarks again

Remarks you typed as `[text]` during counting are stored verbatim at their
position in the grain sequence. `extract_remarks()` pulls them all out across a
site so you can find your way back to whatever you flagged.

```r
extract_remarks(site)
```

Each row gives the sample, the slide, the traverse, the remark, and `id` — the
taxon counted immediately *before* the remark (e.g. `"I1"`), which is what you
look for down the microscope to locate the spot. Pass `id = "after"` for the
grain counted immediately after instead.

```r
# The grain counted just after each remark
extract_remarks(site, id = "after")
```

The `slide` column is the slide that was active when you wrote the remark — the
ID you gave it at the New Slide prompt, or its number within the sample if you
did not name it.

---

## Stratigraphic diagram

`site_matrix()` converts the loaded site into parallel matrices and vectors
ready for `rioja::strat.plot()`. Set `min_present` to hide taxa that appear in
very few samples.

`strat.plot()` plots against a single y-axis, supplied as `yvar`. Pass
`mat$DepTop` for a depth axis or `mat$AgeTop` for an age axis, and set `ylabel`
to match. Axis styling beyond that is rioja's business — see `?rioja::strat.plot`.

```{r site-matrix}
mat <- site_matrix(site, min_present = 2)

cat("Samples:", nrow(mat$TaxaPerc), "\n")
cat("Taxa:   ", ncol(mat$TaxaPerc), "\n")
```

```{r strat-plot, fig.width = 10, fig.height = 7}
if (requireNamespace("rioja", quietly = TRUE)) {
  keep <- colSums(mat$TaxaPerc > 0) >= 3

  rioja::strat.plot(
    mat$TaxaPerc[, keep],
    yvar          = mat$DepTop,
    y.rev         = TRUE,
    scale.percent = TRUE,
    cex.xlabel    = 0.8,
    srt.xlabel    = 45,
    ylabel        = "Depth (cm)",
    title         = "Fake Lake — pollen percentages"
  )
} else {
  message("Install the rioja package to produce the stratigraphic diagram.")
}
```

---

## Concentrations and accumulation rates

`site_matrix()` also returns per-taxon concentrations (`TaxaConc`) and
accumulation rates (`TaxaAccRate`) when spike, sample quantity, and age
interval data are present in the count files.

```{r concentration}
total_conc <- rowSums(mat$TaxaConc, na.rm = TRUE)
data.frame(
  Sample   = rownames(mat$TaxaPerc),
  Depth_cm = mat$DepTop,
  Age_BP   = mat$AgeTop,
  Conc     = round(total_conc)
)
```

Use `accum_rate()` for a dedicated summary that validates all required inputs
up front:

```{r accum-rate}
ar <- accum_rate(site)
print(ar$data)
```

---

## Export to Tilia XML

`write_tlx()` exports the entire site to a `.tlx` file readable by Tilia and
suitable for Neotoma upload. Four spreadsheet pages (Data, Percents,
Concentrations, Accumulation) are written with nine shared metadata header rows
per sample matching Tilia's own output format.

```{r write-tlx}
tlx_path <- tempfile(fileext = ".tlx")
write_tlx(site, file = tlx_path)

cat("File size:", file.size(tlx_path), "bytes\n")

if (requireNamespace("xml2", quietly = TRUE)) {
  doc   <- xml2::read_xml(tlx_path)
  pages <- xml2::xml_find_all(doc, "//SpreadSheet")
  cat("Pages:", paste(xml2::xml_attr(pages, "name"), collapse = ", "), "\n")
}
```

After opening the TLX in Tilia, fill in the *Site* and *Collection Unit* tabs,
add radiocarbon dates on the *Chronology* tab, and run CONISS from the *Analysis*
menu before submitting to Neotoma.

---

## Non-pollen applications

The workflow above applies without modification to other stratigraphic proxies.
Replace the pollen dictionary with one whose codes and groups match your proxy:

- **Diatoms** — code each taxon; use groups to define the sum base (e.g. total
  valves counted).
- **Charcoal morphotypes** — code morphotype categories; accumulation rates
  are computed identically once depths and ages are supplied.
- **Phytoliths, NPPs, dinoflagellates** — any assemblage that is counted by
  traversing a prepared slide fits the model.

The only pollen-specific assumption in the package is the preservation-code
vocabulary (perfect/corroded/crumpled/broken). All other calculations — sums,
concentrations, PAR, matrix output, TLX export — are proxy-agnostic.
